---
title: "Reproducible Research: Peer Assessment 1"
output:
  html_document:
    keep_md: true
---


## 1\. Loading and preprocessing the data

1. Unzip the activity data.
2. Read the csv data from the returned `filepath`.
3. Read the date column as a date type an store it into a day field.
4. Remove the unzipped activity csv file.

```{r}
filepath <- unzip("activity.zip")
data <- read.csv(filepath, stringsAsFactors = F)
data$day <- as.Date(data$date, "%Y-%m-%d")
unlink(filepath)
```

## 2\. What is mean total number of steps taken per day?

#### 2\.1 Histogram of total steps taken per day

1. Load the [dplyr](https://cran.r-project.org/web/packages/dplyr/index.html) package.
2. Group the data by day, and sum the steps by day. Store the result into the variable `steps_by_day`.
3. Plot a histogram of the total number of steps taken per day.

```{r}
library(dplyr)

steps_by_day <-
  group_by(na.omit(data), day) %>%
  summarise(steps = sum(steps))

hist(steps_by_day$steps,
  main = "Histogram steps by day",
  xlab = "distribution (total steps per day)",
  ylab = "frequency (average total days)",
  col = "red")
```

We can see the following from the histogram the following:

- In the first bin we have an average of `5` days where we have an total steps distribution in between `0..5000` steps per day.
- In the middle bin an average of more than `25` days where the distribution lies in between `10000-15000` steps per day.
- In the last bin an average of about `2` days where the distribution lies in between `20000-25000` steps per day.

#### 2\.2 Report the mean and median of total number of steps taken per day.

We summarize the mean and median from the previous `steps_by_day` variable,
and report then.

```{r}
steps_by_day_summary <- summarise(steps_by_day, mean = mean(steps), median = median(steps))
cat(sprintf("Mean:   %0.2f\nMedian: %d",
  steps_by_day_summary$mean,
  steps_by_day_summary$median))
```


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
