---
title: "Reproducible Research: Peer Assessment 1"
output:
  html_document:
    keep_md: true
---


## 1\. Loading and preprocessing the data

1. Unzip the activity data.
2. Read the csv data from the returned `filepath`.
3. Read the date column as a date type an store it into a day field.
4. Remove the unzipped activity csv file.

```{r}
filepath <- unzip("activity.zip")
data <- read.csv(filepath, stringsAsFactors = F)
data$time <- ""
for (i in seq_along(1:nrow(data))) {
  interval <- toString(data$interval[i])
  len      <- nchar(interval)
  if (len > 2) {
    hours   <- as.integer(substr(interval, 0, len - 2))
    minutes <- as.integer(substr(interval, len - 1, len))
  }
  else {
    hours   <- 0
    minutes <- as.integer(interval)
  }

  data$time[i] <- sprintf("%s %02d:%02d:00", data$date[i], hours, minutes)
}
data$date <- as.Date(data$date, "%Y-%m-%d")
data$day  <- format(data$date, "%a")
data$time <- strptime(data$time, format = "%Y-%m-%d %H:%M:%S")
data$time <- as.POSIXct(data$time)
unlink(filepath)
```

## 2\. What is mean total number of steps taken per day?

#### 2\.1 Histogram of total steps taken per day

1. Load the [dplyr](https://cran.r-project.org/web/packages/dplyr/index.html) package.
2. Group the data by day, and sum the steps by day. Store the result into the variable `steps_by_day`.
3. Plot a histogram of the total number of steps taken per day.

```{r}
library(dplyr)

steps_by_day <-
  group_by(na.omit(data), date) %>%
  summarise(steps = sum(steps))

hist(steps_by_day$steps,
  main = "Histogram steps by day",
  xlab = "distribution (total steps per day)",
  ylab = "frequency (average total days)",
  col = "red")
```

We can see the following from the histogram the following:

- In the first bin we have an average of `5` days where we have an total steps distribution in between `0..5000` steps per day.
- In the middle bin an average of more than `25` days where the distribution lies in between `10000-15000` steps per day.
- In the last bin an average of about `2` days where the distribution lies in between `20000-25000` steps per day.

#### 2\.2 Report the mean and median of total number of steps taken per day.

We summarize the mean and median from the previous `steps_by_day` variable,
and report them.

```{r}
steps_by_day_summary <- summarise(steps_by_day,
  mean = mean(steps),
  median = median(steps))

cat(sprintf("Mean:   %0.2f\nMedian: %d",
  steps_by_day_summary$mean,
  steps_by_day_summary$median))
```


## What is the average daily activity pattern?

1. Group the data by interval, and get mean average the steps by interval. Store the result into the variable `steps_by_interval`.
2. Plot a time series over the 5-minute interval and the average steps taken every 5-minute across the days.
3. Sort the `steps_by_interval` by steps in descending order, and display the first row with the max number of steps and its interval.

```{r}
steps_by_interval <-
  group_by(na.omit(data), interval) %>%
  summarise(steps = mean(steps))

with(steps_by_interval,
  plot(x = interval, y = steps,
    type = "l",
    col = "blue",
    main = "Time series Plot",
    xlab = "5-minute intervals over time",
    ylab = "Average steps taken"))

arrange(steps_by_interval, desc(steps)) %>%
  head(1)
```

We can see from the plot the following:

- Activity steps starts usually around 5am in the morning and peaks between 8-9am.
- Then it's steady from 10 am to around 20pm where the activity steps decreases.

The max average steps is at 8:35 in the morning with an average of `206` steps.

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
